---
title: "Bayesian Modelling of ATP Tennis Player Skill from Match Outcomes"
author: "Jensen Suhenda 44377570"
output:
  pdf_document: default
  html_document:
    df_print: paged
urlcolor: cyan
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
library(knitr)
```

[Github Repository](https://github.com/jsuhenda/bayesian-tennis-model.git)

## Dataset
We use a dataset of ATP tennis match results from Kaggle covering professional tennis matches from 2000–2026.

Dataset link:
[ATP Tennis Dataset](https://www.kaggle.com/datasets/dissfya/atp-tennis-2000-2023daily-pull?resource=download&select=atp_tennis.csv)

Each match includes the tournament, date, type of tournament, whether the match was played indoors or not, type of court surface, round of the match, maximum number of sets in the match, participants in the match, winner, ranks of the participants, odds to win and match score.


```{r}
url <- paste0(
  "https://raw.githubusercontent.com/",
  "jsuhenda/bayesian-tennis-model/",
  "refs/heads/main/atp_tennis.csv"
)
df <- read.csv(url)
head(df)
```
\newpage

## Backup Dataset

As a backup dataset, we can use this ATP match results dataset from JeffSackmann's GitHub Repository:
[Backup ATP Tennis Dataset](https://github.com/JeffSackmann/tennis_atp)

This dataset contains historical ATP match results from 1968 to 2024 and includes similar variables:
\begin{itemize}
\item match winner
\item match loser
\item surface
\item tournament
\item rankings
\end{itemize}

The GitHub repository organizes the data by year (e.g., atp_matches_1968.csv contains all match results from 1968). If computational constraints arise when fitting the model to the full dataset, we can restrict the analysis to a subset of years (e.g., 2018–2025) or use a single-year dataset.

## Scientific Questions

The goal of this project is to estimate latent player skill using a Bayesian ranking model.

Our main research questions are:

\begin{itemize}
\item Can a Bayesian latent-skill model recover meaningful rankings of tennis players from match outcomes?
  \begin{itemize}
  \item Using match results, we estimate player skill parameters and evaluate whether the resulting rankings align with
  known top players.
  \end{itemize}

\item Does allowing player skill to vary by court surface improve predictive performance?
  \begin{itemize} 
  \item Some players perform better on certain surfaces (e.g., clay vs grass). We will compare a baseline skill model with   a model that includes surface-specific skill effects.
  \end{itemize}

\item Does allowing player skill to change over time improve match outcome predictions?
  \begin{itemize} 
  \item Player ability evolves over time due to improvement, aging, or injury. We will extend the model to allow player skill to vary across seasons and evaluate whether this dynamic model improves predictive accuracy.
  \end{itemize}
\end{itemize}

The Bayesian approach allows us to estimate posterior distributions for player ability, quantify uncertainty in player rankings, and produce probabilistic predictions for match outcomes.

## Methodology

We model tennis match outcomes using Bayesian latent-skill models based on the Bradley--Terry framework. Each player $i$ is assigned a latent skill parameter $\beta_i$, and the probability that player $i$ defeats player $j$ is modeled as

\[
Y_{ij} \sim \text{Bernoulli}(p_{ij}), \qquad
\text{logit}(p_{ij}) = \beta_i - \beta_j.
\]

This baseline model produces a ranking of players based on match outcomes.

To investigate whether additional structure improves predictions, we consider two extensions. First, we introduce surface-specific skill effects to capture differences in performance across court types. For surface $s$, the model becomes

\[
\text{logit}(p_{ij}) =
(\beta_i + \gamma_{i,s}) -
(\beta_j + \gamma_{j,s}).
\]

Second, we consider a dynamic model where player skill evolves over time. Let $\beta_{i,t}$ denote the skill of player $i$ during time period $t$, and model match outcomes as

\[
\text{logit}(p_{ij,t}) = \beta_{i,t} - \beta_{j,t}.
\]

We will place a temporal structure on the skill parameters, such as a random-walk prior across seasons.

We will also place weakly informative priors on latent skill parameters and any surface or time effects. Since only differences in skill are identifiable, we will impose an identifiability constraint, such as fixing one player’s baseline skill to 0 or constraining the average skill to be 0.

Posterior inference will be performed using Monte Carlo implemented in Stan. Model performance will be evaluated using posterior predictive checks and held-out match prediction to determine whether surface effects and time-varying skill improve predictive accuracy.

### Further Remarks
To isolate the effects of surface and time, these model extensions will be studied separately rather than combined in a single model. A joint model incorporating both surface-specific and time-varying skill parameters would introduce substantially more parameters and computational complexity, and could be something we explore in the future.
