---
title: "Bayesian Modelling of ATP Tennis Player Skill from Match Outcomes"
author: "Jensen Suhenda 44377570"
output: 
  pdf_document:
    latex_engine: xelatex
urlcolor: cyan
geometry: margin=1in
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r libs}
library(dplyr)
library(readr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(knitr)
library(loo)
```

# 1 Introduction

We study the problem of estimating latent tennis player skill from binary match outcomes using Bayesian inference. The core challenge is that player skill is **unobserved** — only win/loss results are recorded, our goal is to infer player ability and evaluate whether additional model structure improves predictive performance.

This is a real-world inference problem in sports analytics, where ranking systems must account for uncertainty in player performance. We model pairwise match outcomes using latent skill parameters and investigate extensions incorporating surface effects and time-varying ability.

Our research questions are:

1. Can a Bayesian latent-skill model recover meaningful rankings from match outcomes?
2. Does allowing surface-specific skill improve predictive performance?
3. Does time-varying skill improve match outcome predictions?

# 2 Data

Dataset link: [ATP Tennis Dataset](https://www.kaggle.com/datasets/dissfya/atp-tennis-2000-2023daily-pull?resource=download&select=atp_tennis.csv)

We use a dataset of ATP tennis matches from 2000–2026, containing approximately 67,000 matches. Each match records the tournament, date, court surface, round, players, match outcome, player rankings, betting odds, and match score.

Under the Bradley--Terry framework, a match reduces to a binary outcome: who won. We therefore retain only the winner, loser, court surface, and year.


```{r, ech0 = FALSE}
df <- read_csv("data/atp_tennis.csv")

df_clean <- df %>%
  mutate(
    winner_name = Winner,
    loser_name = ifelse(Winner == Player_1, Player_2, Player_1),
    surface = Surface,
    date = as.Date(Date),
    year = year(date)
  ) %>%
  select(winner_name, loser_name, surface, year) %>%
  filter(
    !is.na(winner_name),
    !is.na(loser_name),
    !is.na(surface),
    !is.na(date)
  )

head(df_clean)
```

Fitting the full dataset (67,000+ matches across 1,000+ unique players) would result
in an extremely sparse observation matrix: most pairs of players have never met.
Sparse match records lead to weakly identified posterior distributions for most
players. We address this in two ways.

First, we restrict to matches from 2014 onwards, which reduces the data to
approximately the last decade of professional play and aligns with a relatively
stable era of the tour. Second, we retain only the 50 most active players by total
match count. Restricting to well-observed players ensures that each player's
$\beta_i$ is identified by a meaningful number of match outcomes.

Full data wrangling and EDA are in `data_cleaning/wrangling_eda.Rmd` in the [Github
repository](https://github.com/jsuhenda/bayesian-tennis-model.git). The
modelling dataset contains matches among the 50 most active players from
2014–2026.

```{r load-data, include=FALSE}
df_model <- read_csv("data/df_model.csv")

# Player and surface indexing (3 surfaces — Carpet phased out pre-2014)
top_players <- sort(union(df_model$winner_name, df_model$loser_name))
player_ids  <- tibble(player = top_players, player_id = seq_along(top_players))
surface_ids <- tibble(surface = c("Hard", "Clay", "Grass"), surface_id = 1:3)

# 3-year period index for Model 2
df_model <- df_model %>%
  mutate(period = case_when(
    year >= 2014 & year <= 2016 ~ 1,
    year >= 2017 & year <= 2019 ~ 2,
    year >= 2020 & year <= 2022 ~ 3,
    year >= 2023              ~ 4
  ))

# Stan-ready data
df_stan <- df_model %>%
  left_join(player_ids, by = c("winner_name" = "player")) %>%
  rename(winner_id = player_id) %>%
  left_join(player_ids, by = c("loser_name" = "player")) %>%
  rename(loser_id = player_id) %>%
  left_join(surface_ids, by = "surface") %>%
  filter(!is.na(surface_id))

stan_data_m1 <- list(
  N = nrow(df_stan), P = length(top_players), S = 3,
  winner = df_stan$winner_id, loser = df_stan$loser_id,
  surface = df_stan$surface_id
)

stan_data_m2 <- list(
  N = nrow(df_stan), P = length(top_players), T = 4,
  winner = df_stan$winner_id, loser = df_stan$loser_id,
  period = df_stan$period
)
```

# 3 Literature Review

The Bradley--Terry model (Bradley & Terry, 1952) is a standard framework for modeling pairwise comparisons and forms the basis of many sports ranking systems. In this model, the probability that player $i$ defeats player $j$ depends only on the difference in their latent skill parameters, linked through the logistic function.

Extensions of Bradley--Terry to a Bayesian setting have a long history in the sports analytics literature. Glickman (1999) introduced a Bayesian analogue of the Elo rating system in which player strengths are modelled as normally distributed and updated via approximate Bayesian inference after each match. The TrueSkill system (Herbrich et al., 2006) generalised this to multiplayer settings using expectation propagation. 

In tennis specifically, it is well known that player performance depends strongly on court surface. Players often exhibit surface preferences due to differences in playing style and physical conditions. A well-known example is Rafael Nadal, whose performance on clay courts is substantially stronger than on other surfaces. Baker & McHale (2014) applied a paired-comparison model to ATP data and found significant surface-specific components in player skill. This motivates extending the baseline model to include surface-specific skill effects.

# 4 Model Specification

## Baseline Bradley--Terry

Each player $i$ has a latent skill $\beta_i \in \mathbb{R}$. For a match
between players $i$ and $j$:
$$
Y_{ij} \mid \beta \;\sim\; \text{Bernoulli}(p_{ij}), \qquad
\text{logit}(p_{ij}) = \beta_i - \beta_j.
$$
We fix $\beta_1 = 0$ for identifiability and place weakly informative
priors $\beta_i \sim \mathcal{N}(0, 9)$ on all remaining players. The
choice $\sigma = 3$ implies that skill differences of 3 on the logit
scale — corresponding to win probabilities above 95% — are plausible but
not the norm, consistent with the frequency of dominant matchups in
professional tennis.

## Model 1: Surface-Specific Skill

Player $i$'s effective skill on surface $s \in \{\text{Hard, Clay,
Grass}\}$ is $\beta_i + \gamma_{i,s}$, giving:
$$
\text{logit}(p_{ij,s}) = (\beta_i + \gamma_{i,s}) - (\beta_j + \gamma_{j,s}).
$$
We impose $\sum_s \gamma_{i,s} = 0$ per player for identifiability, with
hierarchical prior $\gamma_{i,s} \sim \mathcal{N}(0, \tau_\gamma^2)$,
$\tau_\gamma \sim \text{HalfNormal}(1)$. The shared $\tau_\gamma$
regularises surface effects toward zero when surface data are sparse,
yielding $50 \times 3 = 150$ surface parameters in total.

## Model 2: Time-Varying Skill

To reduce parameter count while capturing career-level trends, we
aggregate seasons into four 3-year periods: 2014--2016, 2017--2019,
2020--2022, and 2023--2026. Player skill evolves across periods $t$ via a
random walk:
$$
\beta_{i,t} = \beta_{i,t-1} + \varepsilon_{i,t}, \quad
\varepsilon_{i,t} \sim \mathcal{N}(0, \tau^2), \quad
\tau \sim \text{HalfNormal}(0.5),
$$
with initial condition $\beta_{i,1} \sim \mathcal{N}(0, 9)$ and reference
player fixed at $\beta_{1,t} = 0$ for all $t$. This gives $49 \times 4 =
196$ skill parameters. Stan code for both models is in the Appendix.

# 5 Posterior Inference

## Method 1: SNIS via simPPLe (correctness check)

We implement the baseline Bradley--Terry model in simPPLe on a small
subset of real matches: the top 5 active players from our dataset and a sample
of 30 matches between them. 

Note that SNIS is impractical on the full dataset: its effective sample size collapses in high dimensions, and simPPLe's non-vectorised R implementation makes large runs prohibitively slow. We therefore apply SNIS to a small toy subset as a correctness check: if SNIS and HMC posterior means agree on this small problem, the Stan implementation is likely correct.

```{r}
# Build a small real-data subset: 5 most active players, 30 sampled matches
toy_players <- df_model %>%
  pivot_longer(c(winner_name, loser_name), values_to = "player") %>%
  count(player, sort = TRUE) %>%
  slice(1:5) %>%
  pull(player)

# put someone at index 1 as reference
ref_player <- "Tsitsipas S."
toy_players <- c(ref_player, setdiff(toy_players, ref_player))
```

```{r, include = FALSE}
toy_df <- df_model %>%
  filter(winner_name %in% toy_players,
         loser_name  %in% toy_players) %>%
  slice_sample(n = 30) %>%
  mutate(
    # convert to index 1..5 within toy_players, then encode as
    # i = lower-indexed player, j = higher-indexed player,
    # outcome = 1 if player i won
    w_idx = match(winner_name, toy_players),
    l_idx = match(loser_name,  toy_players),
    i_idx = pmin(w_idx, l_idx),
    j_idx = pmax(w_idx, l_idx),
    toy_result = as.integer(w_idx == i_idx)
  )

i_idx       <- toy_df$i_idx
j_idx       <- toy_df$j_idx
toy_results <- toy_df$toy_result
```

```{r snis, echo=TRUE, eval=FALSE}
source("utilities/simple.R") 
source("utilities/simple_utils.R")
set.seed(405)

# simPPLe model: player 1 fixed at beta = 0, infer beta_2..beta_5
bt_toy <- function() {
  beta <- c(0, simulate(Norm(0, 3)), simulate(Norm(0, 3)),
               simulate(Norm(0, 3)), simulate(Norm(0, 3)))
  for (k in seq_along(toy_results))
    observe(toy_results[k], Bern(plogis(beta[i_idx[k]] - beta[j_idx[k]])))
  return(beta[5])
}
posterior(bt_toy, 10000)
ess(posterior_particles(bt_toy, 10000))
```
Using Tsitsipas as the reference player ($\beta_{\text{Tsitsipas}} = 0$), SNIS recovers posterior mean skills of $+1.05$ for Djokovic and $-1.77$ for Dimitrov. The ordering is directionally correct --- Djokovic was the strongest player on the tour during this period and Dimitrov the weakest of the five. On the probability scale, these values imply Tsitsipas beats Djokovic with probability $1/(1 + e^{1.05}) \approx 0.26$, and Tsitsipas beats Dimitrov with probability $1/(1 + e^{-1.77}) \approx 0.85$, consistent with known head-to-head records. This gives us initial confidence that the Stan implementation, which uses the same Bradley--Terry likelihood, is correct. The effective sample size is $144$ out of $10{,}000$ particles, giving $\text{RMSE} \approx 2/\sqrt{144} \approx 0.17$, so posterior means are reliable to one decimal place --- sufficient for a correctness check but too coarse for publication-quality estimates, motivating the switch to HMC for the full dataset.

## Method 2: HMC in Stan

We fit both models using Stan's NUTS sampler (Hamiltonian Monte Carlo): 4 chains, 1,000 warmup + 2,000 sampling iterations (8,000 posterior draws).

```{r stan-fit, echo=TRUE, eval=FALSE}
library(cmdstanr)
mod_m1 <- cmdstan_model("stan/model1.stan")
mod_m2 <- cmdstan_model("stan/model2.stan")

fit_m1 <- mod_m1$sample(
  data = stan_data_m1, seed = 405, chains = 4, 
  iter_warmup = 1000, iter_sampling = 2000, adapt_delta = 0.95)

fit_m2 <- mod_m2$sample(
  data = stan_data_m2, seed = 405, chains = 4, 
  iter_warmup = 1000, iter_sampling = 2000, adapt_delta = 0.95)

fit_m1$save_object("fits/fit_m1.rds")
fit_m2$save_object("fits/fit_m2.rds")
```


```{r load-fits, include=FALSE}
# Load the cached fits on every knit
fit_m1 <- readRDS("fits/fit_m1.rds")
fit_m2 <- readRDS("fits/fit_m2.rds")
```

# 6 Results

## Posterior Skill Ranking Plot
```{r, echo=FALSE, fig.width=6, fig.height=3}
beta_df <- fit_m1$summary("beta", mean = mean, ~quantile(.x, c(0.1, 0.9))) %>%
  mutate(player_id = as.integer(gsub("beta\\[|\\]", "", variable))) %>%
  left_join(player_ids, by = "player_id") %>%
  arrange(desc(mean)) %>%
  slice(1:20)

ggplot(beta_df, aes(x = reorder(player, mean), y = mean,
                    ymin = `10%`, ymax = `90%`)) +
  geom_pointrange(size = 0.3) +
  coord_flip() +
  labs(x = NULL, y = "Posterior skill (logit scale)") +
  theme_bw(base_size = 9)
```

## Surface Effects Plot

```{r, echo=FALSE, fig.width=6, fig.height=3}
gamma_df <- fit_m1$summary("gamma", mean = mean, ~quantile(.x, c(0.1, 0.9))) %>%
  mutate(
    idx     = gsub("gamma\\[|\\]", "", variable),
    player_id = as.integer(sub(",.*", "", idx)),
    surf_id   = as.integer(sub(".*,", "", idx))
  ) %>%
  left_join(player_ids, by = "player_id") %>%
  left_join(surface_ids, by = c("surf_id" = "surface_id")) %>%
  filter(player %in% c("Nadal R.", "Djokovic N.", "Alcaraz C.",
                        "Sinner J.", "Federer R."))

ggplot(gamma_df, aes(x = surface, y = mean, ymin = `10%`, ymax = `90%`,
                     color = surface)) +
  geom_pointrange(show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~player, ncol = 5) +
  labs(x = NULL, y = "Surface effect (logit)") +
  theme_bw(base_size = 8)
```

## Career Trajectories Plot
```{r, echo=FALSE, fig.width=6, fig.height=3}
period_labels <- c("2014-16", "2017-19", "2020-22", "2023-26")

beta_t_df <- fit_m2$summary("beta", mean = mean, ~quantile(.x, c(0.1, 0.9))) %>%
  mutate(
    idx       = gsub("beta\\[|\\]", "", variable),
    player_id = as.integer(sub(",.*", "", idx)),
    period    = as.integer(sub(".*,", "", idx))
  ) %>%
  left_join(player_ids, by = "player_id") %>%
  filter(player %in% c("Nadal R.", "Federer R.", "Djokovic N.", "Alcaraz C.", "Sinner J."))

ggplot(beta_t_df, aes(x = period, y = mean, ymin = `10%`, ymax = `90%`,
                      color = player, fill = player)) +
  geom_ribbon(alpha = 0.15, color = NA) +
  geom_line() + geom_point() +
  scale_x_continuous(breaks = 1:4, labels = period_labels) +
  labs(x = NULL, y = "Skill (logit)") +
  theme_bw(base_size = 9)
```

## Model Comparison: Calibration

We compare the two models using a **reliability diagram**, which plots predicted probabilities against observed frequencies across binned probability ranges. A well-calibrated model produces points along the diagonal: when the model says an event has probability $p$, the event actually occurs with frequency $p$.

Since our `generated quantities` block reports the probability that the winner won each match (always 1 by construction), we randomly reframe half the matches from the loser's perspective ($p \to 1-p$, outcome $\to 0$). This spreads predictions across the full 0--1 range with genuine 0/1 outcomes, allowing a meaningful calibration check.

```{r calibration, echo=TRUE, eval=FALSE}
p_winner_m1 <- fit_m1$summary("p_pred")$mean
p_winner_m2 <- fit_m2$summary("p_pred")$mean

set.seed(405)
flip <- rbinom(length(p_winner_m1), 1, 0.5)

calib_df <- bind_rows(
  tibble(p_pred  = ifelse(flip == 1, 1 - p_winner_m1, p_winner_m1),
         outcome = ifelse(flip == 1, 0, 1),
         model   = "Model 1 (surface)"),
  tibble(p_pred  = ifelse(flip == 1, 1 - p_winner_m2, p_winner_m2),
         outcome = ifelse(flip == 1, 0, 1),
         model   = "Model 2 (time)")
) %>%
  mutate(bin = cut(p_pred, breaks = seq(0, 1, 0.1), include.lowest = TRUE)) %>%
  group_by(model, bin) %>%
  summarise(mean_pred = mean(p_pred),
            mean_obs  = mean(outcome),
            n         = dplyr::n(),
            .groups   = "drop")

ggplot(calib_df, aes(x = mean_pred, y = mean_obs, color = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_line() +
  geom_point(aes(size = n), alpha = 0.7) +
  scale_size_continuous(range = c(1, 4), guide = "none") +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Predicted probability", y = "Observed frequency", color = NULL) +
  theme_bw(base_size = 9)
```

Both models are well-calibrated: predicted probabilities closely track observed frequencies across the full 0-1 range, with both curves hugging the diagonal. The two models are nearly indistinguishable in calibration, suggesting that while surface effects and time-varying skill capture different aspects of player performance, neither meaningfully improves calibration over the other. This supports the validity of our Bayesian Bradley-Terry framework: the posterior predicted probabilities are well-behaved and can be interpreted at face value.








\newpage

# Appendix: Stan Code {.unnumbered}

## Model 1 --- Surface Effects {.unnumbered}

```stan
data {
  int<lower=1> N; 
  int<lower=1> P; 
  int<lower=1> S;
  array[N] int winner; 
  array[N] int loser; 
  array[N] int surface;
}
parameters {
  vector[P-1] beta_raw;
  matrix[P, S] gamma_raw;
  real<lower=0> tau_gamma;
}
transformed parameters {
  vector[P] beta;
  beta[1] = 0; beta[2:P] = beta_raw;
  matrix[P, S] gamma;
  for (p in 1:P) gamma[p,] = gamma_raw[p,] - mean(gamma_raw[p,]);
}
model {
  beta_raw ~ normal(0, 3);
  tau_gamma ~ normal(0, 1);
  to_vector(gamma_raw) ~ normal(0, tau_gamma);
  for (n in 1:N) {
    real lp = beta[winner[n]] + gamma[winner[n], surface[n]]
            - beta[loser[n]]  - gamma[loser[n],  surface[n]];
    1 ~ bernoulli_logit(lp);
  }
}
generated quantities {
  vector[N] log_lik;
  vector[N] p_pred;
  for (n in 1:N) {
    real lp = beta[winner[n]] + gamma[winner[n], surface[n]]
            - beta[loser[n]]  - gamma[loser[n],  surface[n]];
    log_lik[n] = bernoulli_logit_lpmf(1 | lp);
    p_pred[n]  = inv_logit(lp);
  }
}
```

## Model 2 --- Time-Varying Skill {.unnumbered}

```stan
data {
  int<lower=1> N; 
  int<lower=1> P; 
  int<lower=1> T;
  array[N] int winner; 
  array[N] int loser; 
  array[N] int period;
}
parameters {
  matrix[P-1, T] beta_raw;
  real<lower=0> tau;
}
transformed parameters {
  matrix[P, T] beta;
  beta[1,] = rep_row_vector(0, T);
  beta[2:P,] = beta_raw;
}
model {
  tau ~ normal(0, 0.5);
  beta_raw[,1] ~ normal(0, 3);
  for (t in 2:T) beta_raw[,t] ~ normal(beta_raw[,t-1], tau);
  for (n in 1:N) {
    1 ~ bernoulli_logit(beta[winner[n], period[n]] - beta[loser[n], period[n]]);
  }
}
generated quantities {
  vector[N] log_lik;
  vector[N] p_pred;
  for (n in 1:N) {
    real lp = beta[winner[n], period[n]] - beta[loser[n], period[n]];
    log_lik[n] = bernoulli_logit_lpmf(1 | lp);
    p_pred[n]  = inv_logit(lp);
  }
}
```


# References {.unnumbered}

Baker, R. D., & McHale, I. G. (2014). A dynamic paired comparisons
model. *European J. Operational Research*, 236(2), 677–684.

Bradley, R. A., & Terry, M. E. (1952). Rank analysis of incomplete block
designs. *Biometrika*, 39(3–4), 324–345.

Glickman, M. E. (1999). Parameter estimation in large dynamic paired
comparison experiments. *J. Royal Statistical Society C*, 48(3),
377–394.

Herbrich, R., Minka, T., & Graepel, T. (2006). TrueSkill: A Bayesian
skill rating system. *NeurIPS*, 19.

